Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-07-07 analysis/paper/paper.Rmd
Minor edits to text and added color to Slinkey plot

```{r libraries, eval = TRUE, results = 'hide'}
library(tidyverse)
library(GGally)
library(ggridges)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(networkD3)
library(webshot)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

@@

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016; @Bicho2020; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181]. In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].

@@

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades. Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g. @solheim2017c, with references]. A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007]. Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @eigeland2015; @fuglestvedt2007; @damlien2016; @solheim2020]. However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].

@@

```{r supp1, cache = FALSE}
radiocarbon %>% 
  select(-excavation_project) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    material = "Material",
    lab_code = "Lab code",
    c14_bp = "C14-age",
    error = "Error"
  )
```

@@

```{r supp2, cache = FALSE}
links <- data.frame(value = 1, categories = names(sites %>% select(core:flake_retouch_other))) %>% 
  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point", "transverse_point", 
                                   "single_edged_point", "hognipen_point", "projectile_undefined"), "projectile",
                          ifelse(categories %in% c("scraper", "knife", "burin", "drill", "blade_retouch",  
                                                     "microblade_retouch", "chip_retouch", "fragment_retouch", 
                                                   "flake_retouch"), "small_tool",
                                 ifelse(categories %in% c("axe", "chisel"), "macro_tool", 
                                 ifelse(categories %in% c("axe_other", "chisel_other", "club_hatchet_other"), "macro_tool_other", 
                                 ifelse(categories %in% c("knife_other", "scraper_other", "drill_other", 
                                                            "microblade_retouch_other", "blade_retouch_other", "fragment_retouch_other",
                                                            "flake_retouch_other"), "small_tool_other", paste0(categories, " ")))))))

nodes <- data.frame(
  name=c(as.character(links$categories), 
         as.character(links$new_categories)) %>% unique()
)

links$IDsource <- match(links$categories, nodes$name)-1 
links$IDtarget <- match(links$new_categories, nodes$name)-1 
nodes$group <- as.factor(c("none"))

my_color <- 'd3.scaleOrdinal() .domain([none"]) .range(["white"])'

sn <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              NodeID = "name", Value = "value",
              sinksRight=TRUE, nodeWidth = 0, fontSize = 14)

htmlwidgets::saveWidget(sn, here::here('analysis/figures/sn1.html'))
webshot::webshot(here::here('analysis/figures/sn1.html'), here::here('analysis/figures/sn1.pdf'), delay = 0.2)
```

>>>>

```{r libraries, eval = TRUE, results = 'hide'}
library(tidyverse)
library(GGally)
library(ggridges)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(networkD3)
library(sf)
library(rnaturalearth)
library(oxcAAR)
quickSetupOxcal()
```

@@

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021] --- the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181]. In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].

@@

As most of the variation in the data is accounted for by the dominating non-flint material in later assemblages, this suppresses and makes it difficult to discern patterns in the flint data. A second CA was therefore run excluding the non-flint material (Figure \@ref(fig:flintcor)). While not as substantial, there is clear temporal patterning in the flint data as well. This is most marked for the very earliest sites which are pulled away from the main cluster, as projectiles, microburins, macro tools, debitage from their production, and flakes characterises these sites. Slightly younger sites appear more impacted by core fragments and blades. The temporal transition in the main cluster is not as marked, but clearly present, and is driven by a larger proportion of blades, flakes and small tools in the earliest assemblages of the cluster, which is opposed to chips, fragments and partly micro-blades. Apart from the impact of core fragments, which is not always highlighted, this must be considered very much in agreement with previous research [e.g. @solheim2017c, with references]. A marked presence of core fragments has, however, previously been noted as one of several similarities between Early Mesolithic Norwegian sites and Late Palaeolithic sites from continental Europe [@fuglestvedt2007]. Overall, the comparatively limited impact of the flint material can possibly be the result of the aggregation of categories that leads to an suppression of otherwise temporally distinct patterns, and there are certainly technological nuances in the flint material that is temporally contingent but not recorded during a regular classification of the material [e.g. @damlien2016; @eigeland2015; @fuglestvedt2007; @solheim2020]. However, while the former pertains to the analytical trade-off between robustness and sensitivity, the latter is likely to be true for the non-flint material as well [@eigeland2007]. The overall pattern does speak to the impact the properties of the raw-material has for the general composition of the assemblages [cf. @manninen2014].

@@

```{r supp1, cache = FALSE}
# Supplementary table of radiocarbon dates
radiocarbon %>% 
  select(-excavation_project) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    material = "Material",
    lab_code = "Lab code",
    c14_bp = "C14-age",
    error = "Error"
  )
```

@@

```{r supp2, cache = FALSE}
# Most of the code for assembling the Sankey plot has been taken directly from 
# various walkthroughs at https://www.r-graph-gallery.com

# Create two columns: old and new variable names
links <- data.frame(value = 1,
        categories = names(sites %>% select(core:flake_retouch_other))) %>% 
  mutate(new_categories = ifelse(categories %in% c("microlith", "tanged_point",
                                 "transverse_point", "single_edged_point", 
                                 "hognipen_point", "projectile_undefined"), 
                                 "projectile",
                          ifelse(categories %in% c("scraper", "knife", "burin",
                                 "drill", "blade_retouch", "microblade_retouch",
                                 "chip_retouch", "fragment_retouch", 
                                 "flake_retouch"), "small_tool",
                          ifelse(categories %in% c("axe", "chisel"),
                                 "macro_tool", 
                          ifelse(categories %in% c("axe_other", "chisel_other",
                                 "club_hatchet_other"), "macro_tool_other", 
                          ifelse(categories %in% c("knife_other", 
                                 "scraper_other", "drill_other", 
                                 "microblade_retouch_other", 
                                 "blade_retouch_other", 
                                 "fragment_retouch_other",
                                 "flake_retouch_other"), "small_tool_other",
                                 paste0(categories, " ")))))))

# Set up nodes and links for the plot
nodes <- data.frame(
  name=c(as.character(links$categories), 
         as.character(links$new_categories)) %>% unique()
)

links$IDsource <- match(links$categories, nodes$name)-1 
links$IDtarget <- match(links$new_categories, nodes$name)-1 
nodes$group <- as.factor(c("none"))
links$group <- as.factor(links$new_categories)

# Specify colours for the plot (this can be assembled using a combination of 
# e.g. unique(links$new_connection), rep() and paste(), but I couldn't be 
# bothered so I just copy+pasted and set it up manually).
my_colour <-'d3.scaleOrdinal() .domain(["core", "core_fragment", "blade", "microblade", "chip", "fragment", "flake", "axe_debitage", "macro_tool", "projectile", "microburin", "small_tool", "beach_flint_flint_nodule", "macro_tool_other", "axe_debitage_other", "grinding_slab_stone", "projectile_other", "small_tool_other", "hammerstone", "flake_other", "fragment_other", "chip_other", "blade_other", "microblade_other", "core_other", "core_fragment_other,"]) .range(["lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "#ff8086", "#79d49f", "lightgrey", "#8eb8fa", "lightgrey", "#a88c7d", "lightgrey", "lightgrey", "lightgrey", "#ffee8c",  "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey", "lightgrey"])'

# Create plot
sn <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              NodeID = "name", Value = "value", colourScale = my_colour,
              sinksRight = FALSE, LinkGroup = "new_categories", 
              NodeGroup="group", nodeWidth = 12, nodePadding = 7, fontSize = 10)

# This has to be saved as html widget and then retrieved as a pdf or png using
# webshot.
htmlwidgets::saveWidget(sn, here::here('analysis/figures/sn1.html'))
webshot::webshot(here::here('analysis/figures/sn1.html'), 
                 here::here('analysis/figures/sn1.pdf'), delay = 0.2)
```
